---
title: "Community Analysis"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

- [P2P community analysis - Google Docs](https://docs.google.com/document/d/1Iu6seHxPXPGH-kiHr2NV0p0hBC_laOgaK1u19hKJWEA/edit#heading=h.v84ckom8081g)
    - [NMDS Tutorial in R – sample(ECOLOGY)](https://jonlefcheck.net/2012/10/24/nmds-tutorial-in-r/)
    - [Lab 8. Communities](http://benbestphd.com/landscape-ecology-labs/lab8.html)

## load packages

```{r}
library(tidyverse)
library(vegan)
library(vegan3d)
library(readxl)
library(DT)

set.seed(2) # for consistent vegan::metaMDS() results
```

## Gregorio's Argentina rocky shore data

- [P2P data templates](https://marinebon.github.io/p2p/methods_data_science.html)
- Gregorio enters data:
  `Biodiv_surverys/Argentina/GBigatti/Data_Sheet_RockyShore_GB_Puerto Madryn.xlsx`
- ↳ Eduardo's P2P transformation:
  [RS_dataTransform](https://marinebon.github.io/p2p-dashboard/RS_dataTransform.html)
- ↳ P2P cleaned data: ~`sites.csv`, ~`obs.csv`
- ↳ P2P analytical products, eg [p2p-dashboard](https://marinebon.github.io/p2p-dashboard), ....

### load data

```{r}
# paths
data_dir  <- "~/Gdrive Ecoquants/projects/mbon/p2p/data/rocky"
sites_csv <- file.path(data_dir, "AR-PMAD_dashboard-siteDF.csv") 
obs_csv   <- file.path(data_dir, "AR-PMAD_dashboard-occurrence.csv") 

# read data
sites      <- read_csv(sites_csv)
obs        <- read_csv(obs_csv)
# View(obs)

datatable(head(obs))
```

### TODO: filter by date, strata

```{r}
# TODO: option to constrain by date range
range(obs$eventDate)

# TODO: option to constrain by strata
obs$strata = factor(obs$strata, levels = c("lowtide", "midtide", "hightide"))
table(obs$strata)
```

### setup abundance data for NMDS

```{r}
# d_n: dataframe(d) of abundance(n)
d_n <- obs %>% 
  filter(
    !is.na(abundance)) %>% 
  group_by(siteCode, scientificName) %>% 
  summarize(
    n = mean(abundance)) %>% 
  ungroup() %>% 
  complete(siteCode, scientificName, fill = list(n = 0)) %>% 
  spread(scientificName, n) %>% 
  column_to_rownames("siteCode")
#View(d_n)

datatable(head(d_n))
```

### run NMDS for abundance data

```{r}
nmds_n <- metaMDS(
  d_n, # site-by-species matrix
  k=2) # number of reduced dimensions

plot(nmds_n, type='t')
```

### setup percent cover data for NMDS

```{r}
# d_pct: dataframe(d) of percent cover(pct)
d_pct <- obs %>% 
  filter(
    !is.na(Cover)) %>% 
  group_by(siteCode, scientificName) %>% 
  summarize(
    pct = mean(Cover)) %>% 
  ungroup() %>% 
  complete(siteCode, scientificName, fill = list(pct = 0)) %>% 
  spread(scientificName, pct) %>% 
  column_to_rownames("siteCode")
#View(d_pct)

datatable(head(d_pct))
```

### run NMDS for percent cover data

```{r}
nmds_pct <- metaMDS(
  d_pct,      # site-by-species matrix
  k=2,        # number of reduced dimensions
  trymax=100) # alleviate non-convergence

stressplot(nmds_pct)
plot(nmds_pct)

ordiplot(nmds_pct, type="n")
orditorp(nmds_pct, display="species", col="red", air=0.01)
orditorp(nmds_pct, display=  "sites", cex= 1.25, air=0.01)
```

### cluster NMDS for percent cover data

```{r}
# transformed Bray-Curtis as amount (density) of species
d_pct_bray_transformed = vegdist(wisconsin(sqrt(d_pct)), method='bray')

# hierarchically cluster sites
clu = hclust(d_pct_bray_transformed, 'average')
plot(clu)

# cutoff clusters to 2 groups
plot(clu)
rect.hclust(clu, k=2) # or by height of similarity h=0.5

grp = cutree(clu, k=2)

plot(nmds_pct, display='sites')
ordihull(nmds_pct, grp, lty=2, col='red', label=T)
```

## Gil's Israel rocky shore data

```{r}
obs_xlsx <- "~/Gdrive Ecoquants/meetings/2019-04 MBON P2P in Mexico/Rilov_Israel rocky_shore_monitoring data_2009-18_for Ben.xlsx"
```

### load and prep data

```{r}
obs_isr    <- read_excel(obs_xlsx, n_max = 8445)
method_isr <- read_excel(obs_xlsx, n_max = 8447, col_types="text") %>% tail(1)

cols_isr <- method_isr %>% 
  gather("column", "method")
i_cols_meta_end <- min(which(cols_isr$method %in% c("#","%")))-1
cols_meta       <- names(obs_isr)[1:i_cols_meta_end]
cols_abundance  <- cols_isr %>% 
  filter(method == "#") %>% 
  pull(column)
cols_pctcover  <- cols_isr %>% 
  filter(method == "%") %>% 
  pull(column)
  
table(obs_isr$Year, obs_isr$Season)
  #      Fall Spring Summer Winter
  # 2009  555      0     60      0
  # 2010  465    240    150    240
  # 2011  450    195    225    240
  # 2012  450    240    225    240
  # 2013  450    120    180    225
  # 2014  450    210    105    120
  # 2015  300    210    120    120
  # 2016  330    120    120     90
  # 2017  120    120    120    240
  # 2018  270    120     90    120

d_n_isr <- obs_isr %>% 
  filter(Year == 2018, Season == "Fall") %>% 
  #select(Trans_Type) %>% table() %>% 
  # Center     Edge Mid-high  Mid-mid 
  #    105       90       45       30
  filter(Trans_Type == "Center") %>% 
  select(Site_name, Unique_Code, !!!cols_abundance) %>% 
  gather("species", "n", -Site_name, -Unique_Code) %>% 
  filter(!is.na(n)) %>% 
  complete(
    species, nesting(Site_name, Unique_Code), 
    fill = list(n = 0)) %>% 
  group_by(
    Site_name, species) %>% 
  summarize(
    n = mean(n)) %>% 
  ungroup() %>% 
  spread(species, n) %>% 
  column_to_rownames("Site_name")
# View(d_n_isr)
  
datatable(head(d_n_isr))
  
d_pct_isr <- obs_isr %>% 
  filter(Year == 2018, Season == "Fall") %>% 
  #select(Trans_Type) %>% table() %>% 
  # Center     Edge Mid-high  Mid-mid 
  #    105       90       45       30
  filter(Trans_Type == "Center") %>% 
  select(Site_name, Unique_Code, !!!cols_pctcover) %>% 
  gather("species", "pct", -Site_name, -Unique_Code) %>% 
  filter(!is.na(pct)) %>% 
  complete(
    species, nesting(Site_name, Unique_Code), 
    fill = list(pct = 0)) %>% 
  group_by(
    Site_name, species) %>% 
  summarize(
    pct = mean(pct)) %>% 
  ungroup() %>% 
  spread(species, pct) %>% 
  column_to_rownames("Site_name")
# View(d_pct_isr)

# View(obs_isr) # tail(obs_isr$Unique_Code)
# View(cols_isr)
# View(method_isr)

datatable(head(d_pct_isr))
```

### run NMDS for percent cover data

```{r}
isr_nmds_pct <- metaMDS(
  d_pct_isr,  # site-by-species matrix
  k=2,        # number of reduced dimensions
  trymax=100) # alleviate non-convergence

stressplot(isr_nmds_pct)
plot(isr_nmds_pct)

ordiplot(isr_nmds_pct, type="n")
orditorp(isr_nmds_pct, display="species", col="red", air=0.01)
orditorp(isr_nmds_pct, display=  "sites", cex= 1.25, air=0.01)

# transformed Bray-Curtis as amount (density) of species
isr_d_pct_bray_transformed = vegdist(wisconsin(sqrt(d_pct_isr)), method='bray')

# hierarchically cluster sites
isr_clu = hclust(isr_d_pct_bray_transformed, 'average')
plot(isr_clu)

# cutoff clusters to 2 groups
plot(isr_clu)
rect.hclust(isr_clu, k=4) # or by height of similarity h=0.5

isr_grp = cutree(isr_clu, k=4)

plot(isr_nmds_pct, display='sites')
ordihull(isr_nmds_pct, isr_grp, lty=2, col='red', label=T)
```
